---
title: "Diamond Analysis Project"
author: "Amoo Samuel"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
data("diamonds")
```

## Introduction 
This project explores the diamonds dataset available in the ggplot2 package. The aim is to uncover insights about how various diamond attributes — such as carat, cut, color, and clarity — influence the price.

## Dataset Summary
I start by examining a summary of the diamonds dataset to get a sense of its structure, variable types, and basic statistics.

## Price Distribution
I examine the distribution of diamond prices to understand pricing trends and outliers.

```{r Price Distribution}
ggplot(diamonds, aes(x = price)) +
  geom_histogram(fill = "steelblue", color = "black", bins = 50) +
  labs(title = "Histogram of Diamond Prices", x = "Price", y = "Count")
```

## Carat vs Price by Cut
This plot visualizes how carat affects price, and how this relationship varies across different cut qualities.

```{r Catat vs Price by Cut}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.4) +
  facet_wrap(~cut) +
  labs(title = "Carat vs Price Faceted by Cut",
       x = "Carat",
       y = "Price") +
  theme_minimal()
```

## Price by Cut
A boxplot helps compare the price distributions across different cut categories.

```{r Price by Cut}
ggplot(diamonds, aes(x = cut, y = price, fill = cut)) +
  geom_boxplot() +
  labs(title = "Price Distribution by Cut", x = "Cut", y = "Price") +
  theme_minimal()
```


## Conclusion 

- Diamond price increases with carat size, as expected.

- Cut quality influences the price significantly.

- There are clear differences in pricing across cut categories, visible in both the scatter and box plots.

This analysis provides useful insights for buyers, sellers, or anyone interested in understanding what drives diamond pricing.

